Supplement: Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
fixed a typo in an equation
changed a section heading to be more consistent
Josh introduction

Viewed as a probabilistic inference problem,
our goal is to jointly infer a single DSL (written $\mathcal{D}$)
and many programs (written $p$), one for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.

@@

To perform inference, we take inspiration from the Helmholtz machine and the Wake/Sleep algorithm~\cite{hinton1995wake},
which alternates between ``waking'', which draws samples from a recognition model (our $q$);
and  ``sleeping'', which trains a recognition model on samples from the generative model (our $\mathcal{D},\theta$).
For us, a \textbf{Wake} cycle means using the recognition model to search for programs that solve the tasks.
A sleep cycle means  updating these models so that the DSL and  recognition model
more closely match the distribution of tasks.
An important difference from the Helmholtz machine
is that our setup has two distinct sleep phases: a phase where we update the generative model
$(\mathcal{D},\theta)$, which we call \textbf{Sleep-G},
and a second phase where we update the recognition model $q$,
 called \textbf{Sleep-R}.

@@

Our goal  is to induce a DSL while finding programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs; (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation is that these steps bootstrap off each other (Fig.~\ref{feeding}):
\\\noindent \textbf{Wake: Searching for programs.}  Our program search  is informed by both the DSL and the recognition model.
When these improve, we find more programs solving the tasks.
\\\noindent\textbf{Sleep-R: Learning a recognition model.} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce a DSL from the programs we found which solve the tasks.
  As we solve more tasks, we can hone in on richer DSLs that more closely match the domain.

@@

Our lower bound $\lowerBound$ is unconventional,
and one might wonder why we do not instead maximize an ELBO-style bound like in a VAE or in the EM algorithm.
We explain in the Supplement that maximizing an ELBO-style bound
leads to a pathological behavior that  causes the model to easily become trapped in local optima.

@@

We want a recognition model that closely approximates the true posteriors over programs. We formulate this as minimizing the expected KL-divergence, $  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]\|q(p|x) \right) \right]$,
or equivalently maximizing
\begin{equation*}
  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
\end{equation*}
where the expectation is taken over tasks. One could take this expectation
over the observed empirical distribution of tasks,
like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
We found it useful to maximize both an autoencoder-style objective $\mathcal{L}_{\text{AE}}$ and a Helmholtz-style objective $\mathcal{L}_{\text{HM}}$, giving the  objective for a recognition model, $\mathcal{L}_{\text{RM}}$:
\begin{align}
\mathcal{L}_{\text{RM}}& = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}\\
\mathcal{L}_{\text{HM}}& = \expect_{p\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right],\text{ $p$ evaluates to $x$}\nonumber\\
\mathcal{L}_{\text{AE}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
  \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p|\mathcal{D},\theta \right]}\log q(p|x)\right]\nonumber
\end{align}
The $\mathcal{L}_{\text{HM}}$ objective is essential for data efficiency:
all of our experiments train \system on only a few hundred tasks, which is too little for
a high-capacity neural network $q$.
Once we bootstrap a $(\mathcal{D},\theta)$,
we can draw unlimited samples from $(\mathcal{D},\theta)$
and train $q$ on those samples.

@@

\subsection{Sleep-G: Learning a DSL (Generative Model)}\label{grammarInductionSection}

>>>>

Viewed as a probabilistic inference problem,
our goal is to jointly infer a single DSL (written $\mathcal{D}$)
and many programs (written $p$), one for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.
 The neural recognition model and the
generative model embodied in the DSL jointly train each other, as they
iteratively learn to solve more programming tasks.

@@

%% To perform inference, we take inspiration from the Helmholtz machine
%% and the Wake/Sleep algorithm~\cite{hinton1995wake}, which alternates
%% between ``waking'', which draws samples from a recognition model (our
%% $q$); and ``sleeping'', which trains a recognition model on samples
%% from the generative model (our $\mathcal{D},\theta$).  For us, a
%% \textbf{Wake} cycle means using the recognition model to search for
%% programs that solve the tasks.  A sleep cycle means updating these
%% models so that the DSL and recognition model more closely match the
%% distribution of tasks.  An important difference from the Helmholtz
%% machine is that our setup has two distinct sleep phases: a phase where
%% we update the generative model $(\mathcal{D},\theta)$, which we call
%% \textbf{Sleep-G}, and a second phase where we update the recognition
%% model $q$, called \textbf{Sleep-R}.

@@

 Our goal is to induce a DSL while finding programs solving each of the
tasks.  We take inspiration from the Wake/Sleep
algorithm~\cite{hinton1995wake}, as well as the
Exploration-Compression algorithm for bootstrap
learning~\cite{Dechter:2013:BLV:2540128.2540316}.  Wake/Sleep
alternates between training a generative model on samples from a
recognition model (our $q$), and training a recognition model on
samples from the generative model (our $\mathcal{D},\theta$).  EC
alternates between exploring the space of solutions to a set of tasks,
and compressing those solutions to suggest new search primitives
for the next exploration stage.
We combine these ideas into an inference strategy that iterates
through three steps: a \textbf{Wake} cycle uses the current DSL and
recognition model to search for programs that solve the tasks.  The
\textbf{Sleep-G} and \textbf{Sleep-R} cycles update the DSL and the
recognition model, respectively.  Crucially, these steps bootstrap 
each off other (Fig.~\ref{feeding}):

\\\noindent \textbf{Wake: Searching for programs.}  Our program search
is informed by both the DSL and the recognition model. When these
improve, we find more programs solving the tasks.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce the DSL
from the programs found in the wake phase, aiming
to maximally compress (or, raise the prior probability of) these
programs.  As we solve more tasks, we hone in on richer DSLs that
more closely match the domain.
\\\noindent \textbf{Sleep-R: Learning a neural recognition model.}  We
update the recognition model training on two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found by the search procedure during waking. As the DSL
improves and as search finds more programs, the recognition model gets
both more data to train on, and better data.

%% Our goal  is to induce a DSL while finding programs solving each of the tasks.
%% Our strategy is to iterate through three steps: (1) searching for programs; (2) learning a neural recognition model -- which we use
%% to accelerate the search over programs -- and (3) improving the DSL.
%% The key observation is that these steps bootstrap off each other (Fig.~\ref{feeding}):
%% \\\noindent \textbf{Wake: Searching for programs.}  Our program search  is informed by both the DSL and the recognition model.
%% When these improve, we find more programs solving the tasks.
%% \\\noindent\textbf{Sleep-R: Learning a recognition model.} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.

@@

@@

We want a recognition model that closely approximates the true posteriors over programs. We formulate this as minimizing the expected KL-divergence, $  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]\|q(p|x) \right) \right]$,
or equivalently maximizing
\begin{equation*}
  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
\end{equation*}
where the expectation is taken over tasks. One could take this expectation
over the observed empirical distribution of tasks,
like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
We found it useful to maximize both an autoencoder-style objective $\mathcal{L}_{\text{AE}}$ and a Helmholtz-style objective $\mathcal{L}_{\text{HM}}$, giving the  objective for a recognition model, $\mathcal{L}_{\text{RM}}$:
\begin{align}
\mathcal{L}_{\text{RM}}& = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}\\
\mathcal{L}_{\text{HM}}& = \expect_{p\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right],\text{ $p$ evaluates to $x$}\nonumber\\
\mathcal{L}_{\text{AE}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
  \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p'|\mathcal{D},\theta \right]}\log q(p|x)\right]\nonumber
\end{align}
The $\mathcal{L}_{\text{HM}}$ objective is essential for data efficiency:
all of our experiments train \system on only a few hundred tasks, which is too little for
a high-capacity neural network $q$.
Once we bootstrap a $(\mathcal{D},\theta)$,
we can draw unlimited samples from $(\mathcal{D},\theta)$
and train $q$ on those samples.

@@

\subsection{Sleep-G: Learning a Generative Model (a DSL)}\label{grammarInductionSection}

### 2018-02-09 supplement.tex
remove some redundant tex
change listings font size
Josh introduction

>>>>

Our lower bound $\lowerBound$ is unconventional,
and one might wonder why we do not instead maximize an ELBO-style bound like in a VAE or in the EM algorithm.
We explain in the Supplement that maximizing an ELBO-style bound
leads to a pathological behavior that  causes the model to easily become trapped in local optima.
